---
title: "Explore Albacore Logbook Data"
format: html
toc: true
toc_float: true
toc-title: "Navigation"
embed-resources: true
editor: source
knitr:
  opts_chunk:
    echo: FALSE
---

## Purpose

Visualize and explore the non-confidential albacore logbook data, obtained 06 April 2026. Look at total data availability, plot patterns.

```{r,warning=F,message=F}
library(tidyverse)
library(terra)
library(sf)
library(viridis)
library(here)
library(rnaturalearth)

tm <- theme_minimal()+theme(panel.border = element_rect(fill=NA,color='black'))
theme_set(tm)
```

## Basemap

```{r,message=F,warning=F}
sf_use_s2(FALSE)
bg <- ne_countries(scale = "medium", returnclass = "sf") |> st_make_valid()

# Deal with the dateline problem
# Define the two hemispheres as clipping polygons
wh <- st_bbox(c(xmin = -180, ymin = -90, xmax = 0, ymax = 90),
                    crs = 4326) |> st_as_sfc()
eh <- st_bbox(c(xmin = 0, ymin = -90, xmax = 180, ymax = 90),
                    crs = 4326) |> st_as_sfc()

# Clip each hemisphere separately, then recombine and shift
bgw <- st_crop(bg, wh)
bge <- st_crop(bg, eh)

bg <- rbind(bgw, bge) |>
  st_shift_longitude()
bg <- bg[st_is_valid(bg),]
```

## Load data

```{r}
dat <- read_csv(here('data','logbook','monthly_1D_albacore_logbook_nonconfidential_Apr2026.csv')) |> 
  select(-1)
dprop <- read_csv(here('data','logbook','monthly_1D_albacore_logbook_propRemoved_Apr2026.csv')) |> 
  select(-1)
```

## Time series

### Total pixels

Number of (1x1 degree) grid cells with data, over time. This gives a sense of the total number of data points we would have for any analysis by year/month/etc.

```{r}
p_ndata <- dat |> 
  group_by(month) |> 
  summarise(n_pixels=n(),.groups = 'drop') |> 
  ggplot(aes(month,n_pixels))+
  geom_line()+
  geom_point()+
  labs(x="Year-Month",y="Number of Cells")
p_ndata
```

By month across years

```{r}
p_ndatam <- dat |> 
  group_by(month) |> 
  summarise(n_pixels=n(),.groups = 'drop') |> 
  rename(date=month) |> 
  mutate(year=year(date),month=month(date)) |> 
  ggplot(aes(factor(month),n_pixels))+
  geom_boxplot()+
  labs(x="Month",y="Number of Cells")
p_ndatam
```

### Mean CPUE

```{r}
mean_cpue_yr <- dat |> 
  rename(date=month) |> 
  mutate(year=year(date),month=month(date)) |> 
  group_by(year) |> 
  summarise(cpue=mean(mean_logCPUE),sd.cpue=sd(mean_logCPUE),.groups = 'drop') |> 
  mutate(upr=cpue+1.96*sd.cpue,lwr=cpue-1.96*sd.cpue)
mean_cpue_month <- dat |> 
  rename(date=month) |> 
  mutate(year=year(date),month=month(date)) |> 
  group_by(month) |> 
  summarise(cpue=mean(mean_logCPUE),sd.cpue=sd(mean_logCPUE),.groups = 'drop') |> 
  mutate(upr=cpue+1.96*sd.cpue,lwr=cpue-1.96*sd.cpue)
p_cpue_yr <- mean_cpue_yr |> 
  ggplot(aes(year,cpue,ymax=upr,ymin=lwr))+
  geom_pointrange()+
  labs(x="Year",y="CPUE")
p_cpue_mth <- mean_cpue_month |> 
  ggplot(aes(month,cpue,ymax=upr,ymin=lwr))+
  geom_pointrange()+
  labs(x="Month",y="CPUE")
p_cpue_yr
p_cpue_mth
```

### Catch centroids

CPUE-weighted centroid of fishing activity

```{r}
cog_ts <- dat |> 
  # convert longitude to 360 degree
  mutate(lon360=lon_1D%%360) |> 
  group_by(month) |> 
  summarise(meanlon=weighted.mean(lon360,mean_logCPUE),
            meanlat=weighted.mean(lat_1D,mean_logCPUE),.groups='drop')
pcog <- cog_ts |>
  pivot_longer(contains("mean"),names_to="direction",values_to="centroid") |> 
  ggplot(aes(month,centroid))+
  geom_line()+
  facet_wrap(~direction,scales='free')
pcog

pcogm <- cog_ts |>
  pivot_longer(contains("mean"),names_to="direction",values_to="centroid") |> 
  mutate(m=month(month),y=year(month)) |> 
  ggplot(aes(y,centroid))+
  geom_line()+
  facet_grid(m~direction,scales='free')
pcogm
```


## Maps

1x1-degree grid

```{r}
dat_sf <- dat |> st_as_sf(coords=c("lon_1D","lat_1D"),crs=4326,remove=F) |> 
  st_shift_longitude()
bb <- st_bbox(dat_sf)
```

```{r}
# Extent offset by 0.5 so cell centers land on whole-number degrees
r <- rast(
  xmin = 161.5,
  xmax = 242.5,
  ymin = 30.5,
  ymax = 54.5,
  resolution = c(1, 1),
  crs = "EPSG:4326"
)

datr <- map(sort(unique(dat_sf$month)),\(x){
  dsub <- dat_sf |> filter(month==x)
  rasterize(dsub,r,field="mean_logCPUE")
})
nms <- dat |> 
  arrange(month) |> 
  mutate(month_name=month.name[month(month)],year=year(month)) |> 
  unite(lyrn,month_name,year) |> 
  distinct(lyrn) |> 
  pull(lyrn)
nms <- paste(1:length(nms),nms,sep="_")
datr <- sds(datr)
names(datr) <- nms
```


```{r,warning=F}
library(tidyterra)
bgc <- bg |> st_crop(bb)

make_plot_mth <- function(inrast,lyr_name){
  ggplot()+
    geom_spatraster(data=inrast)+
    scale_fill_viridis(na.value=NA,limits=c(0,3))+
    geom_sf(data=bgc)+
    labs(fill="log CPUE",title=lyr_name) 
}
# make_plot_mth(datr[[1]],names(datr)[[1]])
```

Make all plot, then make a gif

```{r}
iwalk(datr,\(x,idx) {
  p <- make_plot_mth(x,idx)
  ggsave(here('plots','logbook cpue',paste0(idx,".png")),p,w=6,h=5)
})
```

Make a gif

```{r}
library(magick)
pngs <- list.files(here('plots','logbook cpue'),pattern = "\\.png$", full.names = TRUE)
ord <- regmatches(pngs, gregexpr("(?<=/)\\d+(?=_)", pngs, perl = TRUE)) |> 
  unlist() |> as.integer()
pngs_ord <- match(sort(ord),ord)
pngs <- pngs[pngs_ord]

frames <- image_read(pngs)

# Combine into a GIF and save
animation <- image_animate(frames, fps = 2)  # adjust fps as needed
image_write(animation, here('plots','logbook cpue',"logbook_cpue.gif"))
```


