---
title: "Explore IATTC Data"
author: "Owen R. Liu"
format: html
editor: visual
---

## Purpose

Explore the [publicly-available](https://www.iattc.org/en-US/Data/Public-domain) IATTC tuna effort data.

```{r,include=F,message=F,echo=F}
library(tidyverse)
library(here)
library(sf)
library(terra)
pt <- theme_minimal()+theme(panel.border = element_rect(color='black',fill=NA))
theme_set(pt)
```

## Import

There are three datasets, for three fishery types: pole-and-line, longline, and purse seine.
```{r}
# longline, metric tons and numbers
lld <- read_csv(here('data','PublicLLTunaBillfishMt.csv'))
# pole and line
pld <- read_csv(here('data','PublicLPTunaFlag.csv'))
# purse seine by flag
psd <- read_csv(here('data','PublicPSTunaFlag.csv'))
psdt <- read_csv(here('data','PublicPSTunaSetType.csv'))

# flag codes
flags <- read_csv(here('data','flag_codes.csv'))
```

## Simple Time Series

Longline

```{r}
# total metric tons over time
llts <- lld |> 
  left_join(flags,by=join_by(Flag==Code)) |> 
  dplyr::select(Year:Hooks,Name,contains("mt")) |> 
  pivot_longer(contains("mt"),names_to="species",values_to="mt") |> 
  mutate(date=make_date(year=Year,month=Month,day=1)) |> 
  group_by(date,Name,species) |> 
  summarise(mt=sum(mt),.groups='drop') |> 
  ggplot(aes(date,mt,color=species))+
  geom_line()+
  facet_wrap(~Name)
# albacore only
lltsalb <- lld |> 
  left_join(flags,by=join_by(Flag==Code)) |> 
  dplyr::select(Year:Hooks,Name,contains("mt")) |> 
  pivot_longer(contains("mt"),names_to="species",values_to="mt") |> 
  filter(species=="ALBmt") |> 
  mutate(date=make_date(year=Year,month=Month,day=1)) |> 
  group_by(date,Name,species) |> 
  summarise(mt=sum(mt),.groups='drop') |> 
  ggplot(aes(date,mt,color=species))+
  geom_line()+
  facet_wrap(~Name)
lltsalb
```

Purse seine
```{r}
pstsalb <- psd |> 
  left_join(flags,by=join_by(Flag==Code)) |> 
  dplyr::select(Year,Month,Name,ALB) |> 
  mutate(date=make_date(year=Year,month=Month,day=1)) |> 
  group_by(date,Name) |> 
  summarise(mt=sum(ALB),.groups='drop') |> 
  ggplot(aes(date,mt))+
  geom_line()+
  facet_wrap(~Name)
pstsalb
```

Pole-and-line
```{r}
pltsalb <- pld |> 
  left_join(flags,by=join_by(Flag==Code)) |> 
  dplyr::select(Year,Month,Name,ALB) |> 
  mutate(date=make_date(year=Year,month=Month,day=1)) |> 
  group_by(date,Name) |> 
  summarise(mt=sum(ALB),.groups='drop') |> 
  ggplot(aes(date,mt))+
  geom_line()+
  facet_wrap(~Name)
pltsalb
```

## Make Spatial Grid

Make a spatial grid to map this information. The longline data is on a 5x5degree grid, while the other datasets are on a 1x1 grid.

```{r}
# 5x5 grid
latmin5 <- min(lld$LatC5)-2.5; latmax5 <- max(lld$LatC5)+2.5
lonmin5 <- min(lld$LonC5)-2.5; lonmax5 <- max(lld$LonC5)+2.5
ext5 <- ext(lonmin5,lonmax5,latmin5,latmax5)
gr5 <- rast(ext=ext5,resolution=5,crs="EPSG:4326")
values(gr5) <- 1:ncell(gr5)

# 1x1 grid
latmin1 <- min(pld$LatC1)-0.5; latmax1 <- max(pld$LatC1)+0.5
lonmin1 <- min(pld$LonC1)-0.5; lonmax1 <- max(pld$LonC1)+0.5
ext1 <- ext(lonmin1,lonmax1,latmin1,latmax1)
gr1 <- rast(ext=ext1,resolution=1,crs="EPSG:4326")
values(gr1) <- 1:ncell(gr1)
plot(gr5);plot(gr1)
```

Using this grid, make a function to generically plot some data

```{r}
plot_eff <- function(dat,year,month,species){
  df <- dat |> 
    filter(Year==year,Month==month) |> 
    dplyr::select(contains("Lon"),contains("Lat"),all_of(species)) |> 
    set_names(c("lon","lat","val"))
  out <- rasterize(df,gr1,values=df$val)
  plot(out)
}
dat=pld;year=1984;month=10;species="ALB"

plot_eff(dat,year,month,species)
```


