---
title: "OIF Model Projections"
format: html
toc: true
toc_float: true
toc-title: "Navigation"
embed-resources: true
editor: source
execute:
  echo: true
  warning: false
  message: false
---

```{r,warning=F,message=F,include=F}
library(tidyverse)
library(terra)
library(sf)
library(sdmTMB)
library(viridis)
library(here)
library(rnaturalearth)
library(sdmTMBextra)

tm <- theme_minimal()+theme(panel.border = element_rect(fill=NA,color='black'))
theme_set(tm)
```

## Purpose

Explore changes in projected albacore fishery CPUE as a result of ocean iron fertilizatino (OIF). Based on fishery logbook and albacore species distribution model data, we designed and built a model to predict logbook CPUE in `fdms w cov.qmd`. Now, we take the selected model and project it under simulated, alternative albacore suitability scenarios resulting from OIF in the North Pacific.

## Load Data

The OIF scenarios have been processed into a list of prediction grids, one for each scenario. See `make_cov_proj_grid.R`

```{r}
pgl <- read_rds(here('data','predgrid_all_OIF_scenarios.rds'))
```

Here's our fitted and selected `sdmTMB` model.

```{r}
m <- read_rds(here('output','alb_selected_model.rds'))
```

## Make projections

```{r}
projs <- imap(pgl,\(x,y) predict(m,newdata=x) |> mutate(scenario=y))
projdf <- bind_rows(projs)
```

## Mean CPUE Differences

What sort of comparisons do we want to make between the various iron fertilization scenarios and the baseline (historical) scenario? We could start by differencing them:

```{r}
# the baseline estimates are from scenario 1
# est is the model prediction (i.e., predicted log CPUE)
base_ests <- projs[[1]] |> select(est:est_rf) |> 
  # exponentiate so the difference is made correctly?
  mutate(across(everything(),exp))
diffs <- imap(projs,\(x,y) {
  scen_ests <- x |> 
    select(est:est_rf) |> 
    mutate(across(everything(),exp))
  # subtract baseline estimates
  out <- scen_ests-base_ests
  out |> 
    rename_with(~paste0("d_",.x),everything()) |> 
    mutate(scenario=y)
})
diffdf <- bind_rows(diffs)

# projections with scenario diffs
scens_df <- projdf |> 
  select(-scenario) |> 
  bind_cols(diffdf) |> 
  mutate(scenario=as.factor(scenario))
```

How big of potential differences are we talking here.

```{r}
# this is across all predicted grid cells and time steps
diffsumm <- scens_df |> 
  mutate(expest=exp(est)) |> 
  mutate(propdiff=d_est/expest) |> 
  group_by(scenario) |> 
  summarise(meandiff=mean(d_est),
            q05=as.numeric(quantile(d_est,0.05)),
            q95=as.numeric(quantile(d_est,0.90)),
            meanprop=mean(propdiff),.groups='drop',
            q05prop=as.numeric(quantile(propdiff,0.05)),
            q95prop=as.numeric(quantile(propdiff,0.90)))
knitr::kable(diffsumm,digits = 4)
```

In general, it seems like the projected CPUE changes are within approximately -20% to 6%, across all time steps and grid cells. Here we can see that in smoothed histograms of the differences. 

```{r}
p_props <- scens_df |> 
  filter(scenario!="1") |> 
  mutate(expest=exp(est)) |> 
  mutate(propdiff=d_est/expest) |> 
  ggplot(aes(propdiff,fill=scenario))+
  geom_density(alpha=0.5)+
  scale_fill_manual(values=viridis_pal()(7))+
  labs(x="Proportional Difference")
p_props
```



```{r}
p_diffs <- scens_df |> 
  filter(scenario!="1") |> 
  ggplot(aes(d_est,fill=scenario))+
  geom_density(alpha=0.5)+
  geom_vline(xintercept=0,linetype=2)+
  scale_fill_manual(values=viridis_pal()(7))+
  labs(x="Difference in Predicted CPUE")
p_diffs

ggsave(here('plots','OIF scenarios CPUE diffs density.png'),p_diffs)
```

```{r}
p_diffs2 <- scens_df |>  
  filter(scenario!=1) |> 
  ggplot(aes(exp(est),d_est,color=scenario))+
  geom_point(size=0.5)+
  facet_wrap(~scenario)+
  scale_color_manual(values=viridis_pal()(7))+
  geom_hline(yintercept=0,linetype=2)+
  labs(x="Base Predicted CPUE",y="Difference")
p_diffs2

ggsave(here('plots','OIF scenarios base vs diff.png'),p_diffs2,w=10)
```

```{r}
pdiffs3 <- scens_df |> 
  filter(scenario!="1") |> 
  ggplot(aes(monthfct,d_est,fill=scenario))+
  geom_boxplot(outliers=F)+
  geom_hline(yintercept=0,linetype=2)+
  facet_wrap(~scenario)+
  scale_fill_manual(values=viridis_pal()(7))+
  labs(x="Month",y="CPUE Difference")
pdiffs3

ggsave(here('plots','CPUE diff month OIF scenarios.png'),pdiffs3,w=10)
```
It sort of seems like across many OIF scenarios, predicted CPUE is lower in June/July and a bit higher in August/September, compared to the baseline scenario.

## Spatial Changes

How did spatial patterns of CPUE change in these different scenarios?

```{r}
quickmap <- function(dat, column = est) {
  ggplot()+
    geom_raster(data=dat, aes(Xkm, Ykm, fill = {{ column }})) +
    coord_fixed() +
    scale_fill_gradient2()+
    theme(panel.background=element_rect(fill='gray80'),
          panel.grid = element_blank())
}
```

Example of differences, for scenario 2 relative to baseline

```{r}
p_s2 <- scens_df |> 
  filter(scenario=="2",monthfct!="November") |> 
  quickmap(d_est)+facet_grid(seasonfct~year)+
  labs(fill=expression(Delta ~ "CPUE"))+
  theme(axis.text=element_blank())
p_s2
ggsave(here('plots','OIF scenario 2.png'),p_s2,w=10)
```

What if we looked at the mean difference and the variance among all scenarios.

```{r}
# calculate mean difference in each location
meandiff_df <- scens_df |> 
  group_by(Xkm,Ykm,year,month,season,seasonfct,monthfct) |> 
  # remove the null scenario
  filter(scenario!="1") |> 
  mutate(expest=exp(est)) |> 
  summarise(meandiff=mean(d_est),
            # relative to predicted CPUE
            propdiff=mean(d_est/expest),
            rangediff = max(d_est)-min(d_est),.groups='drop')
```

```{r}
p_meandiff <- meandiff_df |> 
  quickmap(meandiff)+
  facet_grid(seasonfct~year)+
  theme(axis.text=element_blank())+
  labs(fill=expression(Delta ~ "CPUE"),title="Average difference between OIF scenarios and baseline")
p_rangediff <-meandiff_df |> 
  quickmap(rangediff)+
  scale_fill_viridis(option="C")+
  facet_grid(seasonfct~year)+
  theme(axis.text=element_blank())+
  labs(fill=expression(Delta ~ "CPUE"),title="Range of outcomes across OIF scenarios")

p_propdiff <-meandiff_df |> 
  quickmap(propdiff)+
  facet_grid(seasonfct~year)+
  theme(axis.text=element_blank())+
  labs(fill=expression(Delta ~ "CPUE"),title="Mean proportional change across OIF scenarios")
p_meandiff
p_rangediff
p_propdiff

ggsave(here('plots','OIF scenarios mean.png'),p_meandiff,w=10)
ggsave(here('plots','OIF scenarios range.png'),p_rangediff,w=10)
ggsave(here('plots','OIF scenarios proportions.png'),p_propdiff,w=10)
```


